\vspace{1.0cm} \LARGE \bf MERS-CoV spillover at the camel-human interface

### 2017-08-02 mers-structure.tex
Revise Methods.
Resolves #7.

MultiTypeTree module \citep{vaughan_efficient_2014} was used in BEAUti v2.4.3 \citep{bouckaert_beast_2014} to specify a structured coalescent model with two demes -- humans and camels.
This model estimates 4 parameters: rate of coalescence in human viruses, rate of coalescence in camel viruses, rate of migration from the human deme to the camel deme and rate of migration from the camel deme to the human deme.
Analyses were run on codon position partitioned data with two separate HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models specified for codon positions 1+2 and 3.
A relaxed molecular clock with branch rates drawn from a lognormal distribution \citep{drummond_2006} was used to infer the evolutionary rate from date calibrated tips.
Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
All analyses were run across for 200 million states across ten independent Markov chains (MCMC runs).
%\tbc{What was burn-in? How many steps in the MCMC? How was convergence assessed? How were chains combined?}
Due to the increased complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burn-in, convergence assessed in Tracer v1.6 and states combined using LogCombiner distributed with BEAST v2.4.3 \citep{bouckaert_beast_2014}.
Three chains out of ten did not converge and were discarded altogether.
To confirm that demographic inference was derived from the data rather than the prior we set up an additional ten independent analyses where the mean of the exponential distribution prior for migration rate was set to 10.0.
For these analyses two chains did not converge.

As a secondary analysis, we also considered a scenario where alignments split into two fragments (fragment 1 comprised of positions 1-21000, fragment 2 of positions 21000-onwards), with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
Fragment positions were chosen based on consistent identification of the region around nucleotide 21000 as a probable breakpoint by GARD \citep{pond_gard:_2006} by previous studies into SARS and MERS coronaviruses \citep{hon_evidence_2008,dudas_mers-cov_2016}.
All analyses were set to run for 200 million states, subsampling every 20 000 states.
Chains not converging after 200 million states were discarded.
20\% of the states were discarded as burn-in, convergence assessed with Tracer 1.6 and combined with LogCombiner.
Three chains out of ten did not converge.

\subsection*{Demographic inference of MERS-CoV in the reservoir}
In order to infer the demographic history of MERS-CoV in camels we used the results of structured coalescent analyses to identify introductions of the virus into humans.
The oldest sequence from each cluster introduced into humans was used to represent a random draw from the diversity of MERS-CoV circulating in camels.
These sequences were combined with existing sequence data from camels to give us a dataset with minimal demographic signal coming from epidemiological processes in humans.
Sequences belonging to the outgroup clade where most of MERS-CoV sequences from Egypt fall were removed out of concern that MERS epidemics in the Arabian Peninsula and Egypt are distinct epidemics with relatively poor sampling in the latter.
A flexible skygrid tree prior \citep{gill_2013} was used to recover estimates of relative genetic diversity (N$_{e}\tau$) at 50 evenly spaced grid points across six years, ending at the most recent tip in the tree (2015 August) in BEAST v1.8.4 \citep{drummond_bayesian_2012}.
We set up five independent MCMC chains to run for 500 million states, sampling every 50 000 states.
This analysis suffered from poor convergence, where two chains converged onto one stationary distribution, two to another and the last chain onto a third stationary distribution, with high effective sample sizes.%\lmc{already reported above. Why not remove?} \gdc{Never hurts to repeat.}
Demographic trajectories recovered by the two main stationary distributions are very similar and differences between the two appear to be caused by convergence onto different tree topologies, with onward effects on evolutionary rate estimates and distributions of coalescence times.
This non-convergence effect may have been masked previously by the use of all available MERS-CoV sequences from humans which may have lead MCMC towards one of the multiple stationary distributions.

@@

We extracted the times of camel-to-human introductions from the posterior distribution of multitype trees.
This distribution of introduction times was then discretised as follows: for state  $k = 1, 2, \ldots, L$ in the chain,  $Z_{ijk}$ was $1$ if there as an introduction in month $i$ and year $j$ and $0$ otherwise.
We model the variable $Y_{ij} = \sum_{k = 1}^L Z_{ijk}$ with the hierarchical model:
\begin{align*}
  Y_{ij} &\sim \text{Binomial}(L, \theta_{ij}); \\
  logit(\theta_{ij}) &= \alpha_j + \beta_i; \\
  \beta_i &\sim \text{Normal}(0, \sigma_{m}); \\
  \sigma_{m} &\sim \text{Cauchy}(0, 2.5); \\
  \alpha_j &\sim \text{Normal}(\mu_{y}, \sigma_{y}); \\
  \mu_{y}  &\sim  \text{Normal}(0, 1); \\
  \sigma_{y} &\sim \text{Cauchy}(0, 2.5).
\end{align*}

Odds ratios of introductions can then be computed for each month $i$ as $\text{OR}_i = \exp(\beta_i)$.

>>>>

For our primary analysis, the MultiTypeTree module \citep{vaughan_efficient_2014} of BEAST v2.4.3 \citep{bouckaert_beast_2014} was used to specify a structured coalescent model with two demes -- humans and camels.
This model estimates 4 parameters: rate of coalescence in human viruses, rate of coalescence in camel viruses, rate of migration from the human deme to the camel deme and rate of migration from the camel deme to the human deme.
Analyses were run on codon position partitioned data with two separate HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models specified for codon positions 1+2 and 3.
A relaxed molecular clock with branch rates drawn from a lognormal distribution \citep{drummond_2006} was used to infer the evolutionary rate from date calibrated tips.
Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
All analyses were run for 200 million steps across ten independent Markov chains (MCMC runs) and states were sampled every \tbc{XXX} steps.
Due to the complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burn-in, convergence assessed in Tracer v1.6 and states combined using LogCombiner distributed with BEAST v2.4.3 \citep{bouckaert_beast_2014}.
Three chains out of ten did not converge and were discarded altogether.
This left \tbc{XXX} states on which to base posterior inference.

As a secondary analysis to test robustness to choice of prior, we set up an analysis where we increased the mean of the exponential distribution prior for migration rate to 10.0.
All other parameters were identical to the primary analysis and as before 10 independent MCMC chains were run.
In this case, two chains did not converge.
This left \tbc{XXX} states on which to base posterior inference.

As an additional secondary analysis, we also considered a scenario where alignments were split into two fragments (fragment 1 comprised of positions 1-21000, fragment 2 of positions 21000-onwards), with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
Fragment positions were chosen based on consistent identification of the region around nucleotide 21000 as a probable breakpoint by GARD \citep{pond_gard:_2006} by previous studies into SARS and MERS coronaviruses \citep{hon_evidence_2008,dudas_mers-cov_2016}.
All analyses were set to run for 200 million states, subsampling every 20 000 states.
Chains not converging after 200 million states were discarded.
20\% of the states were discarded as burn-in, convergence assessed with Tracer 1.6 and combined with LogCombiner.
Three chains out of ten did not converge.
This left \tbc{XXX} states on which to base posterior inference.

We extracted the times of camel-to-human introductions from the posterior distribution of multitype trees.
This distribution of introduction times was then discretised as follows: for sample  $k = 1, 2, \ldots, L$ from the posterior,  $Z_{ijk}$ was $1$ if there as an introduction in month $i$ and year $j$ and $0$ otherwise.
We model the variable $Y_{ij} = \sum_{k = 1}^L Z_{ijk}$ with the hierarchical model:
\begin{align*}
  Y_{ij} &\sim \text{Binomial}(L, \theta_{ij}); \\
  logit(\theta_{ij}) &= \alpha_j + \beta_i; \\
  \beta_i &\sim \text{Normal}(0, \sigma_{m}); \\
  \sigma_{m} &\sim \text{Cauchy}(0, 2.5); \\
  \alpha_j &\sim \text{Normal}(\mu_{y}, \sigma_{y}); \\
  \mu_{y}  &\sim  \text{Normal}(0, 1); \\
  \sigma_{y} &\sim \text{Cauchy}(0, 2.5).
\end{align*}

Odds ratios of introductions can then be computed for each month $i$ as $\text{OR}_i = \exp(\beta_i)$.
\tbc{Please spell out what parameters are referring to. In English, what is $\alpha_j$ and what is $\beta_i$? I'm having a hard time figuring out what's going on here.}
\tbc{Can you move `\textit{logit}' to the right hand side of the equation (if not too messy)?}

@@

\subsection*{Demographic inference of MERS-CoV in the reservoir}
In order to infer the demographic history of MERS-CoV in camels we used the results of structured coalescent analyses to identify introductions of the virus into humans.
The oldest sequence from each cluster introduced into humans was used to represent a random draw from the diversity of MERS-CoV circulating in camels.
These sequences were combined with existing sequence data from camels to give us a dataset with minimal demographic signal coming from epidemiological processes in humans.
Sequences belonging to the outgroup clade where most of MERS-CoV sequences from Egypt fall were removed out of concern that MERS epidemics in the Arabian Peninsula and Egypt are distinct epidemics with relatively poor sampling in the latter.
A flexible skygrid tree prior \citep{gill_2013} was used to recover estimates of relative genetic diversity ($N_{e}\tau$) at 50 evenly spaced grid points across six years, ending at the most recent tip in the tree (2015 August) in BEAST v1.8.4 \citep{drummond_bayesian_2012}.
We set up five independent MCMC chains to run for 500 million states, sampling every 50 000 states.
This analysis suffered from poor convergence, where two chains converged onto one stationary distribution, two to another and the last chain onto a third stationary distribution, with high effective sample sizes.
Demographic trajectories recovered by the two main stationary distributions are very similar and differences between the two appear to be caused by convergence onto different tree topologies, with onward effects on evolutionary rate estimates and distributions of coalescence times.
\tbc{This statement seemd a bit strong, as it looked like the $N_{e}\tau$ trajectory was highly similar between these different stationary distributions.}
This non-convergence effect may have been masked previously by the use of all available MERS-CoV sequences from humans which may have lead MCMC towards one of the multiple stationary distributions.
